Preferences for the commodification of pensions in Chile: the rol of intergenerational social mobility

### 2025-11-21 paper/paper 2.tex
up antes de borrar tiempo

\setstretch{1.5}
This document was last modified at 2025-11-17 19:08:04 and it was last
rendered at 2025-11-17 19:08:04.

@@

\$\$

ATT\_\{j,k,j\} = E\big[Y(D = k) \mid O = j, D = k \big] -
E\big[Y(D = j) \mid O = j, D = k \big]

\$\$ \{\#eq-att\}

@@

\caption{\label{fig-interact}Effects of intergenerational occupational
mobility on preferences for pension commodification, by meritocratic
perception. Coefficient plot of origin-specific ATT estimates comparing
mobility versus immobility, estimated from interaction models (T ×
Merit). Points show simple slopes of the mobility effect at low merit
and high merit; bars are 95\% confidence intervals. Estimates from
pooled WLS with doubly robust adjustment, wave fixed effects, and
standard errors clustered by individual.}

@@

Figure~\ref{fig-interact} shows that neither the interaction terms nor
the mobility coefficients reach conventional significance across
trajectories, indicating no detectable moderation of the causal mobility
effect by meritocratic perceptions at the 95\% level. Directionally, in
the left panel, the interaction is positive for Middle→High (\(\beta\) =
0.15, 95\% CI {[}−0.03, 0.33{]}) and Low→High (\(\beta\) = 0.05,
{[}−0.14, 0.24{]}), suggesting, if anything, somewhat larger
pro-commodification effects among high-merit respondents when upwardly
mobile, whereas it is negative for High→Low (\(\beta\) = −0.14,
{[}−0.38, 0.10{]}) and High→Middle (\(\beta\) = −0.03, {[}−0.23,
0.17{]}), showing weaker pro-commodification effects under downward
mobility at high merit. Focusing on the treatment effect at low
meritocracy (the reference category) at the right panel, the mobility
coefficient captures the origin-specific ATT comparing movers to their
immobility counterfactual among low-merit respondents. These effects are
small and imprecise across trajectories: Low→Middle (\(\beta\) = −0.02,
{[}−0.09, 0.04{]}) and Low→High (\(\beta\) = −0.03, {[}−0.10, 0.05{]})
are near zero. Middle→High is positive but not significant (\(\beta\) =
0.07, {[}−0.01, 0.14{]}), Middle→Low is null (\(\beta\) = -0.00,
{[}−0.07, 0.06{]}), and downward from high shows negative, borderline
estimates: High→Middle (\(\beta\) = −0.07, {[}−0.15, 0.00{]}) and
High→Low (\(\beta\) = −0.08, {[}−0.15, 0.00{]}). Substantively, I do not
find evidence that meritocratic perceptions moderate the causal effect
of mobility, although the signs align with the idea that upward mobility
at high merit tilts attitudes toward commodification, and downward
mobility at high merit tilts them away.

\caption{\label{tbl-inttime}Effects of intergenerational occupational
mobility on preferences for pension commodification, by survey wave}

@@

\begin{center}
\scalebox{0.8}{
\begin{tabular}{l c c c c c c}
\hline
 & Low-Middle & Low-High & Middle-High & Middle-Low & High-Middle & High-Low \\
\hline
Intercept                              & $0.18$           & $-0.04$          & $0.28$           & $0.20$           & $0.18$            & $0.15$            \\
                                       & $ [-0.36; 0.73]$ & $ [-0.36; 0.29]$ & $ [-0.01; 0.57]$ & $ [-0.11; 0.52]$ & $ [-0.15;  0.50]$ & $ [-0.27;  0.58]$ \\
Mobility treatment                     & $-0.05$          & $-0.03$          & $0.13^{*}$       & $0.06$           & $-0.19^{*}$       & $-0.19^{*}$       \\
                                       & $ [-0.16; 0.06]$ & $ [-0.15; 0.10]$ & $ [ 0.05; 0.22]$ & $ [-0.03; 0.15]$ & $ [-0.33; -0.05]$ & $ [-0.32; -0.05]$ \\
Wave (Ref.= 2016)                      &                  &                  &                  &                  &                   &                   \\
                                       &                  &                  &                  &                  &                   &                   \\
\quad Wave 2018                        & $0.01$           & $0.01$           & $0.10^{*}$       & $0.09^{*}$       & $-0.04$           & $-0.02$           \\
                                       & $ [-0.10; 0.12]$ & $ [-0.10; 0.12]$ & $ [ 0.03; 0.18]$ & $ [ 0.01; 0.17]$ & $ [-0.15;  0.07]$ & $ [-0.15;  0.10]$ \\
\quad Wave 2023                        & $0.03$           & $0.01$           & $0.19^{*}$       & $0.18^{*}$       & $0.05$            & $0.04$            \\
                                       & $ [-0.09; 0.14]$ & $ [-0.12; 0.13]$ & $ [ 0.10; 0.28]$ & $ [ 0.08; 0.27]$ & $ [-0.09;  0.18]$ & $ [-0.11;  0.19]$ \\
Mobility treatment x Wave (Ref.= 2016) &                  &                  &                  &                  &                   &                   \\
                                       &                  &                  &                  &                  &                   &                   \\
\quad Mobility treatment x Wave 2018   & $0.03$           & $-0.06$          & $-0.04$          & $-0.09$          & $0.11$            & $0.06$            \\
                                       & $ [-0.11; 0.17]$ & $ [-0.21; 0.09]$ & $ [-0.17; 0.09]$ & $ [-0.20; 0.03]$ & $ [-0.04;  0.26]$ & $ [-0.09;  0.22]$ \\
\quad Mobility treatment x Wave 2023   & $0.07$           & $0.10$           & $-0.09$          & $-0.07$          & $0.19^{*}$        & $0.18$            \\
                                       & $ [-0.09; 0.23]$ & $ [-0.08; 0.28]$ & $ [-0.23; 0.06]$ & $ [-0.22; 0.08]$ & $ [ 0.01;  0.38]$ & $ [-0.01;  0.37]$ \\
\hline
Controls                               & Yes              & Yes              & Yes              & Yes              & Yes               & Yes               \\
R$^2$                                  & $0.04$           & $0.05$           & $0.07$           & $0.03$           & $0.06$            & $0.06$            \\
Adj. R$^2$                             & $0.03$           & $0.04$           & $0.06$           & $0.02$           & $0.05$            & $0.05$            \\
Num. obs.                              & $878$            & $745$            & $716$            & $773$            & $912$             & $861$             \\
RMSE                                   & $0.40$           & $0.37$           & $0.39$           & $0.39$           & $0.44$            & $0.42$            \\
N Clusters                             & $497$            & $449$            & $443$            & $470$            & $536$             & $525$             \\
\hline
\multicolumn{7}{l}{\scriptsize{Note: Cells contain regression coefficients with confidence intervals in parentheses. $^*$ Null hypothesis value outside the confidence interval..}}
\end{tabular}
}
\label{table:coefficients}
\end{center}

@@

Table~\ref{tbl-inttime} shows the causal heterogeneity of the mobility
effect across the different waves of the survey. Regarding temporal
heterogeneity, the only statistically significant interaction term is
observed for the High→Middle trajectory in 2023
(\(\beta_{\text{int}} = 0.19\)), 95\% CI {[}0.01, 0.38{]}). However,
when this interaction is added to the baseline effect in 2016 (-0.19),
the marginal effect of High→Middle mobility in 2023 is approximately
zero, and its confidence interval includes the null. Thus, rather than
indicating a reversal toward a positive effect, the evidence suggests
that the negative impact of downward mobility from high backgrounds is
confined to 2016 and attenuates over time. No comparable temporal
changes are observed in the remaining trajectories. Overall, the results
provide little support for strong time-conditional heterogeneity in
mobility effects, beyond the fading of the initially negative
High→Middle effect.

@@

\caption{\label{tbl-complete3}Interactions effects between
intergenerational occupational mobility and perceived meritocracy on
preferences for pension commodification, with covariates and wave fixed
effects.}

@@

\begin{center}
\scalebox{0.7}{
\begin{tabular}{l c c c c c c}
\hline
 & Low-Middle & Low-High & Middle-High & Middle-Low & High-Middle & High-Low \\
\hline
Intercept                                                    & $0.15$           & $-0.04$          & $0.30^{*}$       & $0.21$           & $0.08$           & $0.05$           \\
                                                             & $ [-0.36; 0.67]$ & $ [-0.40; 0.31]$ & $ [ 0.04; 0.56]$ & $ [-0.09; 0.51]$ & $ [-0.25; 0.40]$ & $ [-0.30; 0.41]$ \\
Mobility treatment                                           & $-0.02$          & $-0.03$          & $0.07$           & $-0.00$          & $-0.07$          & $-0.08$          \\
                                                             & $ [-0.09; 0.04]$ & $ [-0.10; 0.05]$ & $ [-0.01; 0.14]$ & $ [-0.07; 0.06]$ & $ [-0.15; 0.00]$ & $ [-0.15; 0.00]$ \\
High meritocracy perception (Ref.= Low)                      & $0.05$           & $0.06$           & $-0.01$          & $0.03$           & $0.17^{*}$       & $0.26^{*}$       \\
                                                             & $ [-0.06; 0.16]$ & $ [-0.07; 0.19]$ & $ [-0.09; 0.08]$ & $ [-0.07; 0.13]$ & $ [ 0.00; 0.33]$ & $ [ 0.04; 0.48]$ \\
Mobility treatment x High meritocracy perception (Ref.= Low) & $0.04$           & $0.05$           & $0.15$           & $0.04$           & $-0.03$          & $-0.14$          \\
                                                             & $ [-0.12; 0.20]$ & $ [-0.14; 0.24]$ & $ [-0.03; 0.33]$ & $ [-0.11; 0.19]$ & $ [-0.23; 0.17]$ & $ [-0.38; 0.10]$ \\
\hline
Controls                                                     & Yes              & Yes              & Yes              & Yes              & Yes              & Yes              \\
R$^2$                                                        & $0.04$           & $0.05$           & $0.08$           & $0.04$           & $0.07$           & $0.10$           \\
Adj. R$^2$                                                   & $0.03$           & $0.04$           & $0.07$           & $0.02$           & $0.06$           & $0.08$           \\
Num. obs.                                                    & $878$            & $745$            & $716$            & $773$            & $912$            & $861$            \\
RMSE                                                         & $0.40$           & $0.37$           & $0.39$           & $0.39$           & $0.44$           & $0.41$           \\
N Clusters                                                   & $497$            & $449$            & $443$            & $470$            & $536$            & $525$            \\
\hline
\multicolumn{7}{l}{\scriptsize{Note: Cells contain regression coefficients with confidence intervals in parentheses. $^*$ Null hypothesis value outside the confidence interval..}}
\end{tabular}
}
\label{table:coefficients}
\end{center}

>>>>

\setstretch{1.5}
This document was last modified at 2025-11-19 13:57:11 and it was last
rendered at 2025-11-19 13:57:11.

@@

\begin{equation}\phantomsection\label{eq-att}{ 
ATT_{j,k,j} = E\big[Y(D = k) \mid O = j, D = k \big] - E\big[Y(D = j) \mid O = j, D = k \big]
}\end{equation}

@@

\caption{\label{fig-interact}Effects of intergenerational occupational
mobility on preferences for pension commodification, by meritocratic
perception. Coefficient plot of origin-specific ATT estimates comparing
mobility versus immobility, estimated from interaction models (T ×
Merit). Points show the marginal effects (simple slopes) of the mobility
treatment at low and high merit; bars are 95\% confidence intervals.
Estimates from pooled WLS with doubly robust adjustment, wave fixed
effects, and standard errors clustered by individual.}

@@

Figure~\ref{fig-interact} displays the origin-specific ATT estimates
from the meritocracy interaction models as marginal effects of the
mobility treatment at low and high meritocratic beliefs. For low-merit
respondents, mobility effects are uniformly small and imprecise across
trajectories: Low→Middle (\(\beta\) = −0.02, 95\% CI {[}−0.09, 0.04{]}),
Low→High (\(\beta\) = −0.03, {[}−0.10, 0.05{]}), Middle→Low (\(\beta\) =
−0.00, {[}−0.07, 0.06{]}), and a modest, non-significant positive
estimate for Middle→High (\(\beta\) = 0.07, {[}−0.01, 0.14{]}), while
downward moves from high origin show borderline negative effects
(High→Middle: \(\beta\) = −0.07, {[}−0.15, 0.00{]}; High→Low: \(\beta\)
= −0.08, {[}−0.15, 0.00{]}). Among high-merit respondents, the pattern
is directionally consistent with meritocratic moderation but
statistically robust only for one trajectory: the ATT for Middle→High at
high merit is positive and significant (\(\beta\) = 0.22, {[}0.05,
0.38{]}), whereas Low→Middle (\(\beta\) = 0.02, {[}−0.13, 0.16{]}),
Low→High (\(\beta\) = 0.02, {[}−0.15, 0.20{]}), High→Middle (\(\beta\) =
−0.10, {[}−0.28, 0.09{]}), and High→Low (\(\beta\) = −0.22, {[}−0.44,
0.00{]}) remain statistically indistinguishable from zero. Overall,
these results provide only limited and trajectory-specific evidence that
meritocratic beliefs moderate the causal effect of intergenerational
occupational mobility on preferences for pension commodification, with a
clearer pro-commodification sign only for upward mobility from middle to
high status among high-merit respondents.

@@

\includegraphics[width=0.9\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-interact-wave-1.pdf}

@@

\caption{\label{fig-interact-wave}Effects of intergenerational
occupational mobility on preferences for pension commodification, by
survey wave. Coefficient plot of origin-specific ATT estimates comparing
mobility versus immobility, estimated from interaction models (T ×
Wave). Points show the marginal effects (simple slopes) of the mobility
treatment in 2016, 2018, and 2023; bars are 95\% confidence intervals.
Estimates from pooled WLS with doubly robust adjustment, wave fixed
effects, and standard errors clustered by individual.}

I find no consistent evidence of heterogeneous causal effects across
survey waves. Figure~\ref{fig-interact-wave} displays the marginal
effects of mobility trajectories on market-justice preferences for
pensions in 2016, 2018, and 2023. Across trajectories, the estimated
ATTs remain small and their confidence intervals largely overlap over
time. Upward mobility from middle origin shows a positive and
statistically significant effect in 2016 (Middle→High, \(\beta\) = 0.13,
95\% CI {[}0.05, 0.22{]}) and downward mobility from high origin
displays negative effects at baseline (High→Middle and High→Low, both
\(\beta\) ≈ −0.19, CIs excluding zero), but these patterns attenuate and
become statistically indistinguishable from zero by 2023. Taken
together, the results suggest that any initial pro-commodification
response to upward mobility and de-commodifying response to downward
mobility are not stable over time, and the causal effect of mobility on
preferences for pension commodification is effectively time-invariant
over the 2016--2023 period.

@@

\caption{\label{tbl-complete3}Marginal effects of mobility by
meritocratic perception (Low vs High Meritocracy).}

@@

\centering
\begin{tabular}{l|c|c|c|c|c}
\hline
Trajectory & Merit & Estimate & Std. Error & CI Low & CI High\\
\hline
Low-Middle & Low Meritocracy & -0.024 & 0.034 & -0.091 & 0.042\\
\hline
Low-Middle & High Meritocracy & 0.015 & 0.072 & -0.127 & 0.157\\
\hline
Low-High & Low Meritocracy & -0.029 & 0.038 & -0.104 & 0.046\\
\hline
Low-High & High Meritocracy & 0.024 & 0.088 & -0.148 & 0.195\\
\hline
Middle-High & Low Meritocracy & 0.066 & 0.038 & -0.009 & 0.140\\
\hline
Middle-High & High Meritocracy & 0.215 & 0.084 & 0.051 & 0.379\\
\hline
Middle-Low & Low Meritocracy & -0.004 & 0.034 & -0.071 & 0.063\\
\hline
Middle-Low & High Meritocracy & 0.038 & 0.068 & -0.095 & 0.171\\
\hline
High-Middle & Low Meritocracy & -0.074 & 0.039 & -0.150 & 0.003\\
\hline
High-Middle & High Meritocracy & -0.100 & 0.094 & -0.285 & 0.085\\
\hline
High-Low & Low Meritocracy & -0.077 & 0.039 & -0.153 & 0.000\\
\hline
High-Low & High Meritocracy & -0.216 & 0.112 & -0.436 & 0.005\\
\hline
\end{tabular}

\caption{\label{tbl-complete4}Marginal effects of mobility by survey
wave (2016, 2018, and 2023).}

\centering
\begin{tabular}{l|c|c|c|c|c}
\hline
Trajectory & Wave & Estimate & Std. Error & CI Low & CI High\\
\hline
Low-Middle & 2016 & -0.048 & 0.055 & -0.156 & 0.060\\
\hline
Low-Middle & 2018 & -0.022 & 0.048 & -0.116 & 0.072\\
\hline
Low-Middle & 2023 & 0.022 & 0.058 & -0.092 & 0.135\\
\hline
Low-High & 2016 & -0.025 & 0.062 & -0.146 & 0.096\\
\hline
Low-High & 2018 & -0.088 & 0.050 & -0.186 & 0.011\\
\hline
Low-High & 2023 & 0.072 & 0.066 & -0.058 & 0.202\\
\hline
Middle-High & 2016 & 0.132 & 0.044 & 0.047 & 0.218\\
\hline
Middle-High & 2018 & 0.091 & 0.051 & -0.008 & 0.190\\
\hline
Middle-High & 2023 & 0.045 & 0.061 & -0.074 & 0.164\\
\hline
Middle-Low & 2016 & 0.061 & 0.044 & -0.025 & 0.147\\
\hline
Middle-Low & 2018 & -0.027 & 0.046 & -0.118 & 0.064\\
\hline
Middle-Low & 2023 & -0.010 & 0.057 & -0.122 & 0.103\\
\hline
High-Middle & 2016 & -0.187 & 0.070 & -0.324 & -0.050\\
\hline
High-Middle & 2018 & -0.078 & 0.050 & -0.175 & 0.019\\
\hline
High-Middle & 2023 & 0.006 & 0.063 & -0.118 & 0.130\\
\hline
High-Low & 2016 & -0.186 & 0.068 & -0.319 & -0.053\\
\hline
High-Low & 2018 & -0.125 & 0.053 & -0.228 & -0.022\\
\hline
High-Low & 2023 & -0.009 & 0.068 & -0.141 & 0.124\\
\hline
\end{tabular}
